Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2012-10-02 Dail_Mad_ext.tex
hhhh

Preliminary DM model runs for several species tended to lead to
estimates of survival that were unrealistically high and recruitment
that were unrealistically low, or the reverse (compared to independent
demographic analyses).  The DM models are able to partition changes in
abundance to survival and recruitment in part by making strong
distributional assumptions. % (Dail and Madsen 2011).
When those assumptions are heavily violated the models may proportion
population growth incorrectly into survival and recruitment, even if
they estimates population growth accurately.

Although partitioning population growth into survival and recruitment
is informative, it is not needed for all applications.  Furthermore, a
simpler model would have several merits: faster running time, fewer
total model combinations, and possibly more realistic estimates.
Therefore, we developed a version of the DM model that estimates and
models population growth directly.  In this model, Eq.~\ref{eq:Nt} is
simplified to:
\begin{equation}
  N_{it} \sim \text{Pois}(\exp(r)N_{it-1})
\label{eq:exp}
\end{equation}
where $r$ represents the instantaneous population growth rate.  This,
like the autoregressive version of the model, is a variant on a simple
density-independent exponential model of population growth.
Density-dependent versions of the model are also possible.  For
example:
\begin{equation}
  N_{it} \sim \text{Pois}(N_{it-1}\exp(r(1-N_{it-1}/K))
\label{eq:rick}
\end{equation}
where $K$ is the stable equilibrium of the population and $r$ is the
instantaneous population growth rate at low population densities, and
both parameters are constrained to be positive.  This is a stochastic version
of the Ricker-logistic population growth model \citep{ricker:1954}.  We also
implemented a modified Gompertz-logistic density-dependent
model \citep{hart_gotelli:2011}:
\begin{equation}
  N_{it} \sim \text{Pois}(N_{it-1}\exp(r(1-\log(N_{it-1}+1)/\log(K+1)))
\label{eq:gomp}
\end{equation}
Here the interpretations of $r$ and $K$ are similar to in the Ricker model.

@@

%The zero inflation factor can also affect
%recruitment or population dynamics (see below).
We also considered a zero-inflated negative binomial model of initial
abundance, but this model did not perform well in preliminary tests.

@@

@@

Differences in observers' ability to see, hear, or identify
birds has long been recognized as a potential source of error
in avian point count surveys such as the BBS
\citep{robbins_etal:1986,diefenbach_etal:2003,sauer_etal:1994auk,alldredge_etal:2007auk,campbell_francis:2011}.
Estimating a separate detection probability for each observer can be
difficult and reduces one's ability to estimate the quantities
of interest.  This problem is compounded by the fact that
observers differ greatly in the number of surveys they have
run (so that many observers' separate detection probabilities
could not be accurately estimated).

@@

To include random observer effects in DM models and these extensions,
Eq.~\ref{eq:p1} can be modified to:
\begin{gather}
n_{ijt} \sim \mathrm{Bin}(N_{it}, p_j) \nonumber \\
\mathrm{logit}(p_j) \sim \mathrm{Normal}(\mu_p, \sigma_p)
\label{eq:pobs}
\end{gather}
where $n_{ijt}$ is the number of bird recorded at site $i$ by observer $j$ in
year $t$, $p_j$ is observer-specific detection probability, $\mu_p$ is the mean
detection probability (on the logit scale), and $\sigma_p$ is the standard
deviation of the random observer effects (also on the logit scale).
We have implemented random observer effects in the Bayesian framework
using program \textbf{JAGS} \citep[version 3.2.0]{plummer:2003}
% (version 3.2.0; Plummer 2003)
with the \textbf{R} (R
Development Core Team 2011) package \texttt{rjags} \citep{plummer:2011}
%(Plummer 2011)
interface.

\section{Forecasting Future Population Size}

@@

>>>>

% Preliminary DM model runs for several species tended to lead to
% estimates of survival that were unrealistically high and recruitment
% that were unrealistically low, or the reverse (compared to independent
% demographic analyses).  The DM models are able to partition changes in
% abundance to survival and recruitment in part by making strong
% distributional assumptions. % (Dail and Madsen 2011).
% When those assumptions are heavily violated the models may proportion
% population growth incorrectly into survival and recruitment, even if
% they estimates population growth accurately.

Although partitioning population growth into survival and recruitment
is ideal in terms of providing a mechanistic description of population
dynamics, it is not always possible using simple count data,
especially when the sites are not closed with respect to immigration
and emigration. %Furthermore, a
%simpler model would have several merits: faster running time, fewer
%total model combinations, and possibly more realistic estimates.
If the mechanistic model must be sacrificed, we suggest replacing it
with classical population growth models that have a long history in
ecology. The simplest of which is the exponential growth
model: $f(N_{it-1}) = N_{it-1}e^r$ where $r$ is the intrinsic rate of
increase. Although demographic and environmental stochasticity in $r$
could be modeled as functions of covariates or random effects, a
simpler way of allow for stocasticity is
to regard $N_{it}$ as a Poisson random variable, simplifying
Eq.~\ref{eq:Nt} to:
\begin{equation}
  N_{it} \sim \text{Pois}(\exp(r)N_{it-1}).
\label{eq:exp}
\end{equation}
This,
like the autoregressive version of the model, is a variant on a simple
density-independent exponential model of population growth.
Density-dependent versions of the model are also possible.  For
example:
\begin{equation}
  N_{it} \sim \text{Pois}(N_{it-1}\exp(r(1-N_{it-1}/K))
\label{eq:rick}
\end{equation}
where $K$ is the stable equilibrium of the population and $r$ is the
instantaneous population growth rate at low population densities, and
both parameters are constrained to be positive.  This is a stochastic version
of the Ricker-logistic population growth model \citep{ricker:1954}.  We also
implemented a modified Gompertz-logistic density-dependent
model \citep{hart_gotelli:2011}:
\begin{equation}
  N_{it} \sim \text{Pois}(N_{it-1}\exp(r(1-\log(N_{it-1}+1)/\log(K+1)))
\label{eq:gomp}
\end{equation}
Here the interpretations of $r$ and $K$ are similar to in the Ricker model.

@@

%The zero inflation factor can also affect
%recruitment or population dynamics (see below).
%We also considered a zero-inflated negative binomial model of initial
%abundance, but this model did not perform well in preliminary tests.

@@

\subsection{Environmental and demographic stochasticity}

Even though the DM model is conceptually very simple, even in its most
stripped down form it includes many random effects (the $N$'s) making
it challenging to compute the likelihood or implement MCMC
algorithms. Nonetheless, in some cases additional random effects may
be of interest. For example, in conventional state-space models,
demographic stochasticity is often modeled as $r_t \sim
\text{Norm}(\mu_r, \sigma_r)$.

@@

Variation in detection probability among observers is another source
of random variation that may need to be accounted for applying the DM
model. For example, differences in observers' ability to see, hear, or identify
birds has long been recognized as a potential source of error
in avian point count surveys such as the BBS
\citep{robbins_etal:1986,diefenbach_etal:2003,sauer_etal:1994auk,alldredge_etal:2007auk,campbell_francis:2011}.
Estimating a separate detection probability for each observer can be
difficult and reduces one's ability to estimate the quantities
of interest.  This problem is compounded by the fact that
observers differ greatly in the number of surveys they have
run (so that many observers' separate detection probabilities
could not be accurately estimated).

@@

To include random observer effects in DM models and these extensions,
Eq.~\ref{eq:p1} can be modified to:
\begin{gather}
X_{ijt} \sim \mathrm{Bin}(N_{it}, p_j) \nonumber \\
\mathrm{logit}(p_j) \sim \mathrm{Normal}(\mu_p, \sigma_p)
\label{eq:pobs}
\end{gather}
where $X_{ijt}$ is the number of bird recorded at site $i$ by observer $j$ in
year $t$, $p_j$ is observer-specific detection probability, $\mu_p$ is the mean
detection probability (on the logit scale), and $\sigma_p$ is the standard
deviation of the random observer effects (also on the logit scale).
We have implemented random observer effects in the Bayesian framework
using program \textbf{JAGS} \citep[version 3.2.0]{plummer:2003}
% (version 3.2.0; Plummer 2003)
with the \textbf{R} package \texttt{rjags} \citep{plummer:2011, R-2012}
interface.

\subsection{Alternative Observation Models}

It would be cool to add something on this, but maybe it's asking too much

\subsection{Forecasting Future Population Size}

Ditto

@@

%%% Figures
